---
title: "Estimating Seasonal Behavior States from Biologging Sensor Data"
subtitle: "Examples from Bearded, Ribbon and Spotted Seals in the Bering Sea"
draft: true
author:
- name: Josh M. London
  affiliation: 1
- name: Devin S. Johnson
  affiliation: 1
- name: Paul B. Conn
  affiliation: 1
- name: Brett T. McClintock
  affiliation: 1
- name: Michael F. Cameron
  affiliation: 1
- name: Peter L. Boveng
  affiliation: 1
address:
- code: 1
  address: Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, United States 
  email: josh.london@noaa.gov
  orcid: orcid.org/0000-0002-3647-5046
date: "`r format(Sys.time(), '%d %B, %Y')`"
disclaimer: >
  The scientific results and conclusions, as well as any views or opinions 
  expressed herein, are those of the author(s) and do not necessarily reflect 
  those of NOAA or the Department of Commerce.
abstract: >
  The seasonal timing of key, annual life history events is an important component of many species' ecology. Seasonal periods important to marine mammals often do not align well with typical labels (i.e., spring, summer, winter, fall). The timing of key life history events is well documented only for species found in accessible rookeries or breeding areas. Our knowledge of seasonal timing for species widely dispersed in inaccessible or remote habitats is poor. Here, we employed data from biologging sensors and new statistical modeling to identify and estimate timing of seasonal states for bearded seals (n=7) captured in Kotzebue Sound, Alaska. These animals provide an initial, small dataset we can work with before exapnding to include ribbon (n=82) and spotted seals (n=46) in future interations. Each of these seals is reliant on the seasonal sea ice for pupping, nursing, breeding and molting and these seasons can be characterized by more time spent hauled out on ice, by changes in dive behavior, and by changes in large-scale movement. We are especially interested in the pupping-breeding-molting season, but also use this approach to identify seasonal structure in the non-breeding period. Seasonal periods were treated as separate behavior states that correspond to a hidden Markov process. Hidden Markov models (HMM) are commonly used to estimate behavior states (e.g., foraging, resting, transit) from telemetry data. Typical HMMs, however, have no temporal memory of state assignments and would likely not capture seasonal level states. To address this, we applied a new approximation to a hidden semi-Markov model and specified the transition matrix for the states to mimic the sequential timing of seasons. Dive and haul-out behavior from biologgers were used to estimate these states. The timing and extent of sea ice in the Bering Sea is predicted to change dramatically over the next 50 years and we anticipate bearded, ribbon, and spotted seals might adjust the timing of these life history events in response to those changes.

output:
  html_document:
    uswebr::noaa_report
---

```{r global-options, include=FALSE}
# set global chunk options
library(knitr)
options(width=80)
opts_chunk$set(dev="png",
               dev.args=list(type="cairo",antialias="none"),
               dpi=192,warning=FALSE, message=FALSE)
```

```{r color-pal, include-FALSE}
library(scales)
cblind_main <- c(blue="#0072B2",
                vermillion="#D55E00",
                bluish_green="#009E73",
                reddish_purple="#CC79A7",
                orange="#E69F00")

cblind_pal <- function() {
  manual_pal(unname(cblind_main))
}

scale_colour_cblind <- function(...) {
  discrete_scale("colour", "cblind", cblind_pal(), ...)
}

scale_color_cblind <- scale_colour_cblind

scale_fill_cblind <- function(...) {
  discrete_scale("fill", "cblind", cblind_pal(), ...)
} 
```


## Introduction

Since 2006, over 128 ribbon and spotted seals have been captured and outfitted with bio-logging devices that measure hourly wet/dry proportions. In addition, seven adult bearded seals were released with bio-logging devices. The critical life history events of these seals are all associated with increased haul-out behavior and changes in dive behavior. Additionally, migratory-like movements are also key indications of seasonal states. Hidden Markov models require a geometrically distributed sojourn time in a given state. Hidden semi-Markov models allow an arbitrary sojourn distribution --- the duration an animal spends in a state can depend on the time it has already spent in that state. O’Connell et al (2010) applied Hidden semi-Markov models to the estrus detection in dairy cows and developed the `mhsmm` library for R to support similar analyses.


## Methods and Analysis

We will start by loading the bearded seal data from the `kotzeb0912` package and the ribbon and spotted seal data from the `pepteldata` package. We will also rely on several other packages for the data munging and analysis. Data within these packages --- `kotzeb0912` and `pepteldata`) --- have been processed through the Wildlife Computers Data Portal and additionally munged for more efficient incorporation into analyses like this.

```{r load-eb-data, include=FALSE}
library(kotzeb0912)
library(pepteldata)
library(sp)
library(rgdal)
library(trip)
library(crawl)
library(argosfilter)
library(parallel)
library(dplyr)
library(lubridate)
```

### Model Seal Movement

The first step will be to load in the location data and run these data through the `crawl` package to model seal movement and estimate locations every 6 hours. The `kotzeb0912` and `pepteldata` have slightly different organizational structures and column headings (will be fixed in the near future) so we need to do some work before merging them together.

```{r merge-tel-data}
data("kotzeb0912_locs")
data("kotzeb0912_depths")
data("kotzeb0912_timelines")
data("peptel_locations")
data("peptel_drytimes")
data("peptel_divedepth")

kotzeb0912_locs <- kotzeb0912_locs %>% 
  dplyr::select(deployid,date_time=unique_posix,instr,
                quality,latitude,longitude,error_radius,
                error_semimajor_axis,
                error_semiminor_axis,
                error_ellipse_orientation) %>% 
  dplyr::filter(instr == "Mk10")  

kotzeb0912_depths <- kotzeb0912_depths %>% 
  dplyr::filter(deployid %in% kotzeb0912_locs$deployid) %>% 
  dplyr::rename(date_time=datadatetime)

kotzeb0912_timelines <- kotzeb0912_timelines %>% 
  dplyr::filter(deployid %in% kotzeb0912_locs$deployid) %>% 
  dplyr::rename(date_time=datadatetime)

peptel_locations <- peptel_locations %>% 
  dplyr::select(deployid,date_time=datadatetime,instr=tagtype,
                quality,latitude,longitude,error_radius,
                error_semimajor_axis = error_semi_major_axis,
                error_semiminor_axis = error_semi_minor_axis,
                error_ellipse_orientation) %>% 
  dplyr::mutate(date_time = lubridate::with_tz(date_time,"UTC")) %>% 
  dplyr::filter(instr %in% c("SPLASH","MK10A","MK10","ST16"),
                !deployid %in% kotzeb0912_locs$deployid)

peptel_ids <- peptel_locations %>% group_by(deployid) %>% 
  filter(n()>30) %>% select(deployid)

peptel_locations <- peptel_locations %>% 
  filter(deployid %in% peptel_ids$deployid)

library(xts)
date_unique <-peptel_locations %>% 
  group_by(deployid) %>%
  do(unique_date = xts::make.time.unique(.$date_time,eps=1)) %>%
  tidyr::unnest(unique_date) %>%
  mutate(unique_posix = as.POSIXct(.$unique_date,origin='1970-01-01 00:00:00',tz='UTC')) %>%
  dplyr::arrange(deployid,unique_posix) %>% 
  dplyr::select(unique_posix) %>% 
  ungroup()

peptel_locations <- peptel_locations %>% arrange(deployid,date_time) %>%
  bind_cols(date_unique) %>% select(-date_time) %>% 
  rename(date_time = unique_posix)

peptel_drytimes <- peptel_drytimes %>% 
  dplyr::filter(deployid %in% peptel_locations$deployid) %>% 
  dplyr::mutate(datadatetime = with_tz(datadatetime,"UTC")) %>% 
  dplyr::select(deployid,date_time=datadatetime,percent_dry=dry)

peptel_divedepth <- peptel_divedepth %>% 
  tidyr::gather(bin,num_dives, starts_with('bin')) %>%
  dplyr::mutate(date_time = with_tz(datadatetime, "UTC")) %>% 
  dplyr::select(deployid,
                instr=tagtype,
                date_time,
                -datadatetime,
                histtype,bin,num_dives) %>%
  dplyr::select(deployid,date_time,num_dives,bin) %>% 
  dplyr::filter(deployid %in% peptel_locations$deployid) %>% 
  dplyr::arrange(deployid,date_time,bin) 

locs <- bind_rows(peptel_locations,kotzeb0912_locs)
depths <- bind_rows(peptel_divedepth,kotzeb0912_depths)
drytimes <- bind_rows(peptel_drytimes,kotzeb0912_timelines)

locs <- locs %>% 
  mutate(species = ifelse(grepl("^EB.",deployid),'Bearded seal',NA),
         species = ifelse(grepl("^PL.",deployid),'Spotted seal',species),
         species = ifelse(grepl("^HF.",deployid),'Ribbon seal',species)) %>% 
  dplyr::filter(!is.na(species)) %>% arrange(deployid,date_time)

depths <- depths %>% 
  mutate(species = ifelse(grepl("^EB.",deployid),'Bearded seal',NA),
         species = ifelse(grepl("^PL.",deployid),'Spotted seal',species),
         species = ifelse(grepl("^HF.",deployid),'Ribbon seal',species))%>% 
  dplyr::filter(!is.na(species)) %>% arrange(deployid,date_time)

drytimes <- drytimes %>% 
  mutate(species = ifelse(grepl("^EB.",deployid),'Bearded seal',NA),
         species = ifelse(grepl("^PL.",deployid),'Spotted seal',species),
         species = ifelse(grepl("^HF.",deployid),'Ribbon seal',species))%>% 
  dplyr::filter(!is.na(species)) %>% arrange(deployid,date_time)

```

Prior to modeling the movement with `crawl`, we will use the `argosfilter` package to pass the Argos locations through a course speed filter (vmax=3.5 m/s).

```{r argos-filter}
# speedfilter using the paralell package for multi-core speed
cfilter<-mclapply(split(locs,locs$deployid),function(x) sdafilter(
  lat=x$latitude, lon=x$longitude, dtime=x$date_time,
  lc=x$quality, ang=-1,vmax=3.5),mc.preschedule=F,mc.cores=3)
cfilter<-do.call("c",cfilter)
cfilter<-as.vector(cfilter)
locs$filtered <- cfilter

locs_data <- filter(locs,filtered=="not") %>% arrange(deployid,date_time) %>% 
  as.data.frame(.)

```

A few of the bearded seals were deployed with GPS capable tags and so we need to merge the GPS data with our Argos data.

```{r gps-data}
data("kotzeb0912_gps")

kotzeb0912_gps <- kotzeb0912_gps %>% 
  mutate(species = "Bearded seal") %>% 
  select(deployid,species,date_time,latitude,longitude) %>% 
  filter(!is.na(latitude)) %>% 
  mutate(error_semimajor_axis = 50,error_semiminor_axis = 50, error_ellipse_orientation = 1) 

locs_data <- locs_data %>% bind_rows(kotzeb0912_gps) %>% 
  filter(!grepl("^EB2004.",deployid)) %>% 
  filter(!grepl("^EB2005.",deployid)) %>% 
  filter(!grepl("^EB2006.",deployid)) %>% 
  filter(!grepl("^PL2005.",deployid)) %>% 
  filter(!grepl("^PL2006.",deployid)) %>% 
  arrange(deployid,date_time)

locs_data_ellipse<- filter(locs_data,!is.na(error_semimajor_axis)) %>% 
  arrange(deployid,date_time) %>% 
  as.data.frame(.)

locs_data_argos <- filter(locs_data,is.na(error_semimajor_axis),
                          !deployid %in% locs_data_ellipse$deployid) %>% 
  arrange(deployid,date_time) %>% 
  as.data.frame(.)
```

The specifications for the movement model are fairly straightforward. Note we are using the development version of `crawl` and incorporating the error ellipse structure. The GPS locations were all given a fixed error of 50m.

```{r crawl-model-ellipse, results='hide'}
data <- as.data.frame(locs_data_ellipse)
coordinates(data) = ~longitude+latitude
proj4string(data) = CRS("+proj=longlat +datum=WGS84")

data=spTransform(data, CRS("+init=epsg:3571"))

## Loop over PTT and fit CTCRW models
ids = unique(data@data$deployid)

library(doParallel)
n.cores <- detectCores()
registerDoParallel(cores=3)

model_fits_ellipse <-
  foreach(i = 1:length(ids)) %dopar% {
  id_data = subset(data,deployid==ids[i])
  diag_data = model.matrix(
      ~ error_semimajor_axis + error_semiminor_axis + error_ellipse_orientation,
      id_data@data
    )[,-1]
    
    id_data@data = cbind(id_data@data, 
                         argosDiag2Cov(
                           diag_data[,1], 
                           diag_data[,2], 
                           diag_data[,3]))
    
    init = list(a = c(coordinates(id_data)[1,1],0,
                      coordinates(id_data)[1,2],0),
                P = diag(c(5000 ^ 2,10 * 3600 ^ 2, 
                           5000 ^ 2, 10 * 3600 ^ 2)))
    
  fit <- crwMLE(
      mov.model =  ~ 1,
      err.model = list(
        x =  ~ ln.sd.x - 1, 
        y =  ~ ln.sd.y - 1, 
        rho =  ~ error.corr
      ),
      data = id_data,
      Time.name = "date_time",
      initial.state = init,
      fixPar = c(1,1,NA,NA),
      theta = c(log(10), 3),
      initialSANN = list(maxit = 2500),
      control = list(REPORT = 10, trace = 1)
    )
    fit
  }

names(model_fits_ellipse) <- ids

data <- as.data.frame(locs_data_argos)
data$quality <- factor(data$quality,
                       levels=c("3", "2", "1","0","A","B"))

coordinates(data) = ~longitude+latitude
proj4string(data) = CRS("+proj=longlat +datum=WGS84")

data=spTransform(data, CRS("+init=epsg:3571"))

## Loop over PTT and fit CTCRW models
ids = unique(data@data$deployid)

library(doParallel)
n.cores <- detectCores()
registerDoParallel(cores=3)

model_fits_argos <-
  foreach(i = 1:length(ids)) %dopar% {
  id_data = subset(data,deployid==ids[i])
    
    init = list(a = c(coordinates(id_data)[1,1],0,
                      coordinates(id_data)[1,2],0),
                P = diag(c(5000 ^ 2,10 * 3600 ^ 2, 
                           5000 ^ 2, 10 * 3600 ^ 2)))
    
    fixPar = c(log(250), log(500), log(1500), rep(NA,4), NA)
    
    constr=list(lower=c(rep(log(1500),2), rep(-Inf,3)),
            upper=rep(Inf,5))
    
    ln.prior = function(theta){-abs(theta[4]-3)/0.5}
    
  fit <- crwMLE(
      mov.model =  ~ 1,
      err.model = list(
        x =  ~ quality - 1
      ),
      data = id_data,
      Time.name = "date_time",
      initial.state = init,
      fixPar = fixPar,
      constr = constr,
      prior = ln.prior,
      initialSANN = list(maxit = 2500),
      control = list(REPORT = 10, trace = 1)
    )
    fit
  }

names(model_fits_argos) <- ids


# print(model_fits_ellipse)
# print(model_fits_argos)
```

### Predict locations every 6 hours

This function predicts the regular-timed -- in this case, 6-hourly -- locations along the movement path using the posterior mean and variance of the track. 

```{r message=FALSE}
predData_ellipse <- foreach(i = 1:length(model_fits_ellipse)) %dopar% {
  if(!inherits(model_fits_ellipse[[i]],"try-error")) {
  model_fits_ellipse[[i]]$data$date_time <- lubridate::with_tz(
    model_fits_ellipse[[i]]$data$date_time,"UTC")
  predTimes <- seq(
    lubridate::ceiling_date(min(model_fits_ellipse[[i]]$data$date_time),"day"),
    lubridate::floor_date(max(model_fits_ellipse[[i]]$data$date_time),"day"),
    "6 hours")
  tmp = crwPredict(model_fits_ellipse[[i]], predTime=predTimes)
  } else return(NA)
}

predData_ellipse <- Filter(function(x) inherits(x,"crwPredict"),predData_ellipse)

predData_ellipse <- dplyr::bind_rows(predData_ellipse) %>% 
  filter(locType=="p")

predData_argos <- foreach(i = 1:length(model_fits_argos)) %dopar% {
  if(!inherits(model_fits_argos[[i]],"try-error")) {
  model_fits_argos[[i]]$data$date_time <- lubridate::with_tz(
    model_fits_argos[[i]]$data$date_time,"UTC")
  predTimes <- seq(
    lubridate::ceiling_date(min(model_fits_argos[[i]]$data$date_time),"day"),
    lubridate::floor_date(max(model_fits_argos[[i]]$data$date_time),"day"),
    "6 hours")
  tmp = crwPredict(model_fits_argos[[i]], predTime=predTimes)
  } else return(NA)
}

predData_argos <- Filter(function(x) inherits(x,"crwPredict"),predData_argos)

predData_argos <- dplyr::bind_rows(predData_argos) %>% 
  filter(locType=="p")

predData <- dplyr::bind_rows(predData_ellipse,predData_argos)

predData$predTimes <- intToPOSIX(predData$TimeNum)

```


### Merge predicted locations with haul-out data

The haul-out behavior timelines are provided as hour percent-dry values. We want to group these values into 6-hour blocks that are centered on our 6-hourly predictions. To accomplish this, we will setup a grouping column that assigns a unique integer to each 6 hour period.

```{r load-timelines, include=FALSE}
drytimes <- drytimes %>% 
  filter(deployid %in% predData$deployid)
```

```{r create-6hr-groups, include=FALSE}
ids <- unique(drytimes$deployid)
timelineData <- foreach(i = 1:length(ids)) %dopar% {
  t_dat <- drytimes %>% 
    filter(deployid == ids[i]) %>% 
    arrange(deployid,date_time)
  t_seq <- data.frame(date_time = seq(
    lubridate::floor_date(min(t_dat$date_time),"day"),
    lubridate::floor_date(max(t_dat$date_time),"day"),
    "1 hour"),deployid=ids[i])
  t_dat <- merge(t_dat,t_seq,all=TRUE)
  rep_len <- ceiling((nrow(t_dat)-3)/6)
  t_group <- c(rep(1,3),rep(2:rep_len,each=6))
  t_group <- data.frame(group=t_group[1:nrow(t_dat)])
  t_dat <-cbind(t_dat,t_group) %>% filter(!is.na(group))
}

timelineData <- dplyr::bind_rows(timelineData) %>% 
  group_by(deployid,group) %>% 
  summarise(percent_dry=mean(percent_dry,na.rm=TRUE),
            date_time=ceiling_date(median(date_time),"hour")) %>% 
  select(deployid,date_time,percent_dry) %>% 
  filter(!is.nan(percent_dry))
```

```{r merge-timeline-with-locs, include=FALSE}
predData <- dplyr::left_join(predData,
                              timelineData,
                              by = c("predTimes" = "date_time", 
                                     "deployid")) %>%
  tbl_df() %>%
  select(deployid,species,predTimes,mu.x,mu.y,percent_dry) %>% 
  arrange(deployid, predTimes) %>% 
  rename(date_time=predTimes)
```

### Merge predicted locations with dive behavior

Dive behavior data are transmitted as dive histograms that represent the distribution of dives across predetermined depth bins for a given 6-hour period. To simplify things, we will sum the number of dives across all bins less than 10m. This should represent the most likely number of foraging dives over a 6 hour period. Future analysis may consider using time-at-depth instead of number of dives.

```{r load-dive-depths}

diveData <- depths %>% 
  filter(bin != "bin1") %>% 
  mutate(date_time = date_time + hours(3)) %>% 
  group_by(deployid,date_time) %>% 
  summarise(num_dives=sum(num_dives,na.rm=TRUE)) %>% 
  ungroup()

predData <- dplyr::left_join(predData,diveData) %>% 
  arrange(deployid,date_time) 
```

### Calculate x and y displacement for each step

We'll use the dplyr::lag() function to calculate the difference in x and y when compared with the previous x and y values. This will give us a measure of northing and easting displacement at each time step. These values will form the basis for our xy-displacement multi-variate normal parameter in the model.

```{r xy-displacement}
predData <- predData %>% 
  group_by(deployid) %>% 
  mutate(x_disp = order_by(date_time, mu.x - lag(mu.x)), 
         y_disp = order_by(date_time, mu.y - lag(mu.y))
  ) %>% 
  filter(!is.na(x_disp) | !is.na(y_disp)) %>% 
  ungroup()
```

We can use a custom function to calculate the compass bearing between those two points. Compass bearing provides a more user friendly description of movement compared to x/y displacement.

```{r compass-bearing}
anglefun <- function(xx,yy,bearing=TRUE,as.deg=FALSE){
  ## calculates the compass bearing of the line between two points
  ## xx and yy are the differences in x and y coordinates between two points
  ## Options:
  ## bearing = FALSE returns +/- pi instead of 0:2*pi
  ## as.deg = TRUE returns degrees instead of radians
  c = 1
  if (as.deg){
    c = 180/pi
  }
  
  b<-sign(xx)
  b[b==0]<-1  #corrects for the fact that sign(0) == 0
  tempangle = b*(yy<0)*pi+atan(xx/yy)
  if(bearing){
    #return a compass bearing 0 to 2pi
    #if bearing==FALSE then a heading (+/- pi) is returned
    tempangle[tempangle<0]<-tempangle[tempangle<0]+2*pi
  }
  return(tempangle*c)
}

predData <- predData %>% rowwise() %>% 
  mutate(bearing = anglefun(x_disp,y_disp,as.deg=TRUE)) %>% 
  ungroup()
```

### Add deploy_day for aligning deployments

Since these deployments occurred over multiple years, we will want a convenient way to align the deployments. The simplest way to do this is to use day-of-year integers and add 365 to days in January-May. At this point, we will also do some additional filtering to make sure we don't have any corrupt or other outlier parameters from the tag data.

```{r days-since-apr1}
id_apr1 <- predData %>% 
  group_by(deployid) %>% summarise(start_yr = year(min(date_time))) %>% 
  mutate(apr1 = lubridate::ymd(paste(start_yr,"04-01",sep="-"))) %>% 
  select(-start_yr) %>% ungroup()

predData <- predData %>% left_join(id_apr1) %>% 
  mutate(days_since_apr1 = as.numeric(difftime(date_time,apr1,units="days")))


predData <- predData %>% 
  mutate(ho_binary=ifelse(percent_dry>=33.3,1,-1),
         y_disp = ifelse(abs(y_disp)>100000,NA,y_disp),
         x_disp = ifelse(abs(x_disp)>100000,NA,x_disp),
         num_dives = ifelse(num_dives > 150,NA,num_dives)) 

predData <- predData %>% 
  filter(days_since_apr1 <=400)

select_ids <- predData %>% group_by(deployid) %>% 
  filter(n()>720) %>% select(deployid)

remove_ids <- c("PL2009_2005_06L0126","HF2005_5898_05L0001","HF2009_1019_09L0040",
                "HF2005_5880_05L0003","HF2005_5882_05L0008","HF2005_5884_05L0009",
                "HF2005_5890_05L0011","HF2005_5894_05L0013","HF2005_5896_05L0015")

predData <- predData %>% 
  filter(deployid %in% select_ids$deployid,
         !deployid %in% remove_ids)

load('pep_age.rda')

predData <- predData %>% left_join(pep_age)

save(predData,file='predData.rda')
```

We now present a series of four plots showing the raw, 'observed' parameter values that will go into the model: haul-out status, number of dives, y-displacement, and x-displacement.

```{r plot-1, fig.height = 8,echo=FALSE}
library(ggplot2)
library(ggthemes)
library(scales)
library(uswebr)

predData_eb <- predData %>% filter(species=="Bearded seal")
p1 <- ggplot(predData_eb, aes(x=days_since_apr1, y=ho_binary)) + 
  geom_point(alpha=0.2,size=3,aes(colour=factor(ho_binary))) + facet_grid(deployid ~ .) + 
  ggtitle("haul-out status (bearded seal)") +
  usweb_theme() + scale_colour_cblind()
p1

predData_hf <- predData %>% filter(species=="Ribbon seal")
p1 <- ggplot(predData_hf, aes(x=days_since_apr1, y=ho_binary)) + 
  geom_point(alpha=0.1,size=2) + facet_grid(deployid ~ .) + 
  ggtitle("haul-out status (ribbon seal)") +
  usweb_theme()+ scale_colour_colorblind()
p1

predData_pl <- predData %>% filter(species=="Spotted seal")
p1 <- ggplot(predData_pl, aes(x=days_since_apr1, y=ho_binary)) + 
  geom_point(alpha=0.1,size=2) + facet_grid(deployid ~ .) + 
  ggtitle("haul-out status (spotted seal)") +
  usweb_theme()+ scale_colour_colorblind()
p1
```

```{r plot-2, fig.height = 8,echo=FALSE}
p2 <- ggplot(predData_eb, aes(x=days_since_apr1, y=num_dives)) + 
  geom_point(alpha=0.2) + facet_grid(deployid ~ .) + 
  ggtitle("number of dives below 10m (bearded seal)") +
  usweb_theme()
p2

p2 <- ggplot(predData_hf, aes(x=days_since_apr1, y=num_dives)) + 
  geom_point(alpha=0.2) + facet_grid(deployid ~ .) + 
  ggtitle("number of dives below 10m (ribbon seal)") +
  usweb_theme()
p2

p2 <- ggplot(predData_pl, aes(x=days_since_apr1, y=num_dives)) + 
  geom_point(alpha=0.2) + facet_grid(deployid ~ .) + 
  ggtitle("number of dives below 10m (spotted seal)") +
  usweb_theme()
p2
```

```{r plot-3, fig.height = 8, echo=FALSE}
predData_eb %>% mutate(direction = ifelse(
  y_disp >= 0, "north","south")) %>% 

ggplot(aes(x=days_since_apr1, y=y_disp/1000)) +
  geom_bar(aes(color=factor(direction)),alpha=0.2,stat="identity") +
  facet_grid(deployid ~ .) + 
  ggtitle("y displacement (bearded seal)") +
  usweb_theme() + scale_colour_cblind()
```

```{r plot-4, fig.height = 8,echo=FALSE}
predData_eb %>% mutate(direction = ifelse(
  y_disp >= 0, "west","east")) %>% 

ggplot(aes(x=days_since_apr1, y=x_disp/1000)) +
  geom_bar(aes(color=factor(direction)),alpha=0.2,stat="identity") +
  facet_grid(deployid ~ .) + 
  ggtitle("x displacement (bearded seal)") +
  usweb_theme() + scale_colour_cblind()
```

### multivariate hidden semi-markov

We are going to use the mhsmm package to run our multi-variate hidden semi-markov model. For the model, we will use 6-hour time steps and the following parameters:

#. haul-out status (Bernouli; cuttoff at 33.3% dry)
#. number of dives below 10m (Poisson)
#. x-y displacement (multi-variate normal)

### bearded seal fit

```{r mshmm-start-eb}
library(mhsmm)

dd <- predData_eb

id = levels(factor(dd$deployid))
dd$ind_dry = ifelse(dd$percent_dry>100/3, 1, 0)
dd$y_disp_km = dd$y_disp/1000
dd$x_disp_km = dd$x_disp/1000

tmp1 = dd#[predData$deployid==id[1],]

### MHSMM functions
source("mhsmm_functions.R")
```

After loading the package and doing a little bit of data cleaning, we need to setup the initial values and parameters. Initial investigation and model comparison with AIC suggests that 4 states provides the best model.

```{r mshmm-init-eb}
# Number of states
J = 4

# Initial vals
init0 = rep(1/J, J)
B0 = list(
  p = c(0.05,0.05,0.07,0.1),
  lambda = rep(30, J), 
  mu = rep(list(c(0,0)), J),
  sigma = rep(list(20*diag(2)), J)
)
P0 = exp(-abs(row(diag(J)) - col(diag(J))))
diag(P0) = 0
P0 = sweep(P0, 1, rowSums(P0), "/")
S0 = list(lambda=rep(150, J), shift=rep(200,J), type="poisson")

start_val = hsmmspec(
  init = init0, 
  transition = P0, 
  parms.emission = B0, 
  sojourn = S0,
  dens.emission = dtelem.hsmm.mv, 
  mstep = mstep.telem.mv
)
```

Now, we can fit the model and predict

```{r mshmm-fit-eb}
# Fit model
data = list(x=with(tmp1, as.matrix(tmp1[,c("ind_dry","num_dives","x_disp_km","y_disp_km")])), N=table(tmp1$deployid))
#M <- 600
fit = hsmmfit(data, start_val, mstep = mstep.telem.mv)

dd$state = predict(fit, data)$s
dd$state = as.factor(dd$state)

fit_bearded <- fit
dd_bearded <- dd
```


### ribbon seal fit

```{r mshmm-start-hf}
library(mhsmm)

dd <- predData_hf %>% filter(age != "YOY")

id = levels(factor(dd$deployid))
dd$ind_dry = ifelse(dd$percent_dry>100/3, 1, 0)
dd$y_disp_km = dd$y_disp/1000
dd$x_disp_km = dd$x_disp/1000

tmp1 = dd#[predData$deployid==id[1],]

### MHSMM functions
source("mhsmm_functions.R")
```

After loading the package and doing a little bit of data cleaning, we need to setup the initial values and parameters. Initial investigation and model comparison with AIC suggests that 4 states provides the best model.

```{r mshmm-init-hf}
# Number of states
J = 3

# Initial vals
init0 = rep(1/J, J)
B0 = list(
  #p = c(0.05,0.05,0.07,0.1),
  p = c(0.1, 0.1, 0.1),
  lambda = rep(30, J), 
  mu = rep(list(c(0,0)), J),
  sigma = rep(list(20*diag(2)), J)
)
P0 = exp(-abs(row(diag(J)) - col(diag(J))))
diag(P0) = 0
P0 = sweep(P0, 1, rowSums(P0), "/")
S0 = list(lambda=rep(150, J), shift=rep(240,J), type="poisson")

start_val = hsmmspec(
  init = init0, 
  transition = P0, 
  parms.emission = B0, 
  sojourn = S0,
  dens.emission = dtelem.hsmm.mv, 
  mstep = mstep.telem.mv
)
```

Now, we can fit the model and predict

```{r mshmm-fit-hf}
# Fit model
data = list(x=with(tmp1, as.matrix(tmp1[,c("ind_dry","num_dives","x_disp_km","y_disp_km")])), N=table(tmp1$deployid))
#M <- 600
fit = hsmmfit(data, start_val, mstep = mstep.telem.mv)

dd$state = predict(fit, data)$s
dd$state = as.factor(dd$state)

fit_ribbon <- fit
dd_ribbon <- dd
```

### YOY spotted seal fit

```{r mshmm-start-pl}
library(mhsmm)

dd <- predData_pl %>% filter(age == "YOY")

id = levels(factor(dd$deployid))
dd$ind_dry = ifelse(dd$percent_dry>100/3, 1, 0)
dd$y_disp_km = dd$y_disp/1000
dd$x_disp_km = dd$x_disp/1000

tmp1 = dd#[predData$deployid==id[1],]

### MHSMM functions
source("mhsmm_functions.R")
```

After loading the package and doing a little bit of data cleaning, we need to setup the initial values and parameters. Initial investigation and model comparison with AIC suggests that 4 states provides the best model.

```{r mshmm-init-pl}
# Number of states
J = 3

# Initial vals
init0 = rep(1/J, J)
B0 = list(
  #p = c(0.05,0.05,0.07,0.1),
  p = c(0.1, 0.1, 0.1),
  lambda = rep(50, J), 
  mu = rep(list(c(0,0)), J),
  sigma = rep(list(20*diag(2)), J)
)
P0 = exp(-abs(row(diag(J)) - col(diag(J))))
diag(P0) = 0
P0 = sweep(P0, 1, rowSums(P0), "/")
S0 = list(lambda=rep(100, J), shift=rep(240,J), type="poisson")

start_val = hsmmspec(
  init = init0, 
  transition = P0, 
  parms.emission = B0, 
  sojourn = S0,
  dens.emission = dtelem.hsmm.mv, 
  mstep = mstep.telem.mv
)
```

Now, we can fit the model and predict

```{r mshmm-fit-pl}
# Fit model
data = list(x=with(tmp1, as.matrix(tmp1[,c("ind_dry","num_dives","x_disp_km","y_disp_km")])), N=table(tmp1$deployid))
#M <- 600
fit = hsmmfit(data, start_val, mstep = mstep.telem.mv)

dd$state = predict(fit, data)$s
dd$state = as.factor(dd$state)

fit_spotted_yoy <- fit
dd_spotted_yoy <- dd
```

## Results

### Adult Bearded Seals

The results of the model fit are provided

```{r fit-summary-eb}
summary(fit_bearded)
```

And, we can plot the state assignments for each bearded seal

```{r plot-5,fig.height=6,echo=FALSE}
library(ggplot2)
ggplot(data=dd_bearded,aes(x=days_since_apr1,y="")) + 
  geom_tile(aes(fill=state)) + 
  facet_wrap(~deployid, ncol=1) + 
  usweb_theme()+
   theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Seasonal State Assignments for 7 Adult Bearded Seals") +
  usweb_theme() + scale_fill_cblind()
```

We can also combine the state assignments across our tagged seals by taking the majority state for each time step. The color transparency is set to the proportion of animals represented by that majority state. Note that, some states may not be present in the combined graph.

```{r plot-6, fig.height=4, echo=FALSE}
combo <- dd_bearded %>% 
  group_by(days_since_apr1,state) %>% 
  summarise(counter = n()) %>% 
  group_by(days_since_apr1) %>% 
  filter(counter == max(counter))

ggplot(data=combo,aes(x=days_since_apr1,y=" ")) + 
  geom_tile(aes(fill=state,alpha=counter/7)) +
  scale_fill_discrete(limits=levels(dd_bearded$state),drop=TRUE) +
  scale_alpha(guide = 'none') + usweb_theme()+
   theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Combined Seasonal State Assignments Across 7 Adult Bearded Seals") + scale_fill_cblind()
```

Now that we have some seasonal states assigned, we can examine the distribution of various behaviors across those different states.

```{r plot-7, echo=FALSE}
ggplot(data=dd_bearded %>% filter(percent_dry>3),aes(x=percent_dry)) +
  geom_density(aes(fill=state),size=0) + facet_wrap(~state) +
  ggtitle("Distribution of Percent Dry Across Seasonal States") +
  usweb_theme() + scale_fill_cblind()
```

```{r plot-8, echo=FALSE}
ggplot(data=dd_bearded %>% filter(num_dives<150),aes(x=num_dives)) +
  geom_density(aes(fill=state),size=0) + facet_wrap(~state) +
  ggtitle("Distribution of Number of Dives Across Seasonal States") +
  usweb_theme() + scale_fill_cblind()
```

```{r plot-9, fig.height=7, echo=FALSE}
ggplot(data=dd_bearded) +
  geom_histogram(aes(x=bearing,y=..density..,fill=state),binwidth=12) +
  coord_polar() + 
  scale_x_continuous(limits=c(0,360),breaks=seq(0,359, by=45)) +
  facet_wrap(~state,ncol=2) +
  ggtitle(paste("Distribution of Movement and \n",
                "Compass Bearing Across Seasonal States")) + 
  usweb_theme()
```


### Adult and Sub-Adult Ribbon Seals

The results of the model fit are provided

```{r fit-summary-hf}
summary(fit_ribbon)
```

And, we can plot the state assignments for each ribbon seal

```{r plot-10,fig.height=6,echo=FALSE}
library(ggplot2)
ggplot(data=dd_ribbon,aes(x=days_since_apr1,y="")) + 
  geom_tile(aes(fill=state)) + 
  facet_wrap(~deployid, ncol=1) + 
  usweb_theme()+
   theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Seasonal State Assignments for Each Ribbon Seal") +
  usweb_theme()
```

We can also combine the state assignments across our tagged seals by taking the majority state for each time step. The color transparency is set to the proportion of animals represented by that majority state. Note that, some states may not be present in the combined graph.

```{r plot-11, fig.height=4, echo=FALSE}
combo <- dd_ribbon %>% 
  group_by(days_since_apr1,state) %>% 
  summarise(counter = n()) %>% 
  group_by(days_since_apr1) %>% 
  filter(counter == max(counter))

ggplot(data=combo,aes(x=days_since_apr1,y=" ")) + 
  geom_tile(aes(fill=state,alpha=counter/7)) +
  scale_fill_discrete(limits=levels(dd_ribbon$state),drop=TRUE) +
  scale_alpha(guide = 'none') + usweb_theme()+
   theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Combined Seasonal State Assignments Across Ribbon Seals")
```

Now that we have some seasonal states assigned, we can examine the distribution of various behaviors across those different states.

```{r plot-12, echo=FALSE}
ggplot(data=dd_ribbon %>% filter(percent_dry>3),aes(x=percent_dry)) +
  geom_density(aes(fill=state),size=0) + facet_wrap(~state) +
  ggtitle("Distribution of Ribbon Seal Percent Dry Across Seasonal States") +
  usweb_theme()
```

```{r plot-13, echo=FALSE}
ggplot(data=dd_ribbon %>% filter(num_dives<150),aes(x=num_dives)) +
  geom_density(aes(fill=state),size=0) + facet_wrap(~state) +
  ggtitle("Distribution of Number of Dives\n for Ribbon Seals Across Seasonal States") +
  usweb_theme()
```

```{r plot-14, fig.height=7, echo=FALSE}
ggplot(data=dd_ribbon) +
  geom_histogram(aes(x=bearing,y=..density..,fill=state),binwidth=12) +
  coord_polar() + 
  scale_x_continuous(limits=c(0,360),breaks=seq(0,359, by=45)) +
  facet_wrap(~state,ncol=2) +
  ggtitle(paste("Distribution of Ribbon Seal Movement and \n",
                "Compass Bearing Across Seasonal States")) + 
  usweb_theme()
```


### Young-of-Year Spotted Seals

The results of the model fit are provided

```{r fit-summary-pl}
summary(fit_spotted_yoy)
```

And, we can plot the state assignments for each young spotted seal

```{r plot-15,fig.height=6,echo=FALSE}
library(ggplot2)
ggplot(data=dd_spotted_yoy,aes(x=days_since_apr1,y="")) + 
  geom_tile(aes(fill=state)) + 
  facet_wrap(~deployid, ncol=1) + 
  usweb_theme()+
   theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Seasonal State Assignments for Each YOY Spotted Seal") +
  usweb_theme()
```

We can also combine the state assignments across our tagged seals by taking the majority state for each time step. The color transparency is set to the proportion of animals represented by that majority state. Note that, some states may not be present in the combined graph.

```{r plot-16, fig.height=4, echo=FALSE}
combo <- dd_spotted_yoy %>% 
  group_by(days_since_apr1,state) %>% 
  summarise(counter = n()) %>% 
  group_by(days_since_apr1) %>% 
  filter(counter == max(counter))

ggplot(data=combo,aes(x=days_since_apr1,y=" ")) + 
  geom_tile(aes(fill=state,alpha=counter/7)) +
  scale_fill_discrete(limits=levels(dd_spotted_yoy$state),drop=TRUE) +
  scale_alpha(guide = 'none') + usweb_theme()+
   theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Combined Seasonal State Assignments Across YOY Spotted Seals")
```

Now that we have some seasonal states assigned, we can examine the distribution of various behaviors across those different states.

```{r plot-17, echo=FALSE}
ggplot(data=dd_spotted_yoy %>% filter(percent_dry>3),aes(x=percent_dry)) +
  geom_density(aes(fill=state),size=0) + facet_wrap(~state) +
  ggtitle("Distribution of Spotted Seal Percent Dry Across Seasonal States") +
  usweb_theme()
```

```{r plot-18, echo=FALSE}
ggplot(data=dd_spotted_yoy %>% filter(num_dives<150),aes(x=num_dives)) +
  geom_density(aes(fill=state),size=0) + facet_wrap(~state) +
  ggtitle("Distribution of Number of Dives\n for Spotted Seals Across Seasonal States") +
  usweb_theme()
```

```{r plot-19, fig.height=7, echo=FALSE}
ggplot(data=dd_spotted_yoy) +
  geom_histogram(aes(x=bearing,y=..density..,fill=state),binwidth=12) +
  coord_polar() + 
  scale_x_continuous(limits=c(0,360),breaks=seq(0,359, by=45)) +
  facet_wrap(~state,ncol=2) +
  ggtitle(paste("Distribution of Spotted Seal Movement and \n",
                "Compass Bearing Across Seasonal States")) + 
  usweb_theme()
```



## Discussion



## References
